Profitability Prediction of Consulting Projects in the Construction Industry

### 2016-08-30 finalwriteup/draft.rmd
finished minor edits

Engaging in loss making jobs for fixed fees is a major problem in consulting, particularly in the competitive construction industry. This thesis investigates whether machine learning techniques applied to a company's passively collected internal data could help avoid loss making jobs or help tactfully choose when to enforce stricter contracts. It was found that in a specific decision framework, a case study's profits could be improved 9% by avoiding approximately 4% of projects. Alternative decision frameworks are also proposed and evaluated. The internal data collected by the case study company described each of their projects in terms of employee positions, fraction of project time completed by each employee, client characteristics, project characteristics (time span, hours, technical details) and invoicing history. Algorithmic methods such as Logistic Regression, Random Forests, Boosted Trees, Naive Bayes, and Bayesian Networks were applied as well as blended combinations of these methods. A decision scenario which" rejected projects above a sequence of tested thresholds was run in order to find the optimal threshold for profit improvements. This process was repeated on hundreds of models built from different training subsets of the data in order to obtain a 95% statistical confidence interval of profit improvements. The blended Logistic Regression model outperformed other methods and produced a 95% confidence interval of 6.5 - 11.5% profit improvements. Further work is recommended to test user input estimation and user interface designs. The findings from this research have the potential to assist managers in reducing losses by highlighting risky projects and guiding project-based changes to fee structures.

@@

**ANOVA** - Analysis of Variance. A statistical test that can be applied to find whether there is a significant difference between the means of more than two groups.

@@

**B2B** - Business to Business. Commerce transactions between two businesses.

**BIM** - Building Information modeling. A digital representation of physical characteristics of an object or structure.

@@

**CLV** - Customer Lifetime Value. Prediction of the profit a customer will bring over the course of a business' relationship with the customer.

**CRM** - Customer Relationship Management. Refers to a common type of software used by businesses to record client and project details as well as employee time sheet records. Generally accessible to each employee in a company.

@@

@@

**ROC** - receiver Operating Characteristic. An ROC curve plots the performance of a binary classification algorithm its scoring threshold is varied.

@@

@@

Predicting how long it will take a person to do something is a notoriously tricky task. People make mental calculations of this sort every day - estimating how long it will take to drive to meet someone on time, how many tasks you can check off in a day, or how long it will take to cook a dish. As most people have experienced, these estimates can be particularly inaccurate if the task is new and difficult. These errors are trivial in everyday activities, but for consulting businesses that solve complex problems, similar time estimation errors determine their financial well-being. Furthermore,

@@

“It’s tough to make predictions, especially about the future” - Source Unknown

@@

Another strategy for managing this risk is by stipulating a specific one-off variation opportunity in the contract. This applies to consulting work that falls under the client's larger project budget. The variation would specify a percentage fee of the overall project cost that can overrule the original fixed price. The purpose of this rule is to guard against the case where the project creeps in size, and the consultant finds themselves gradually undertaking more work and more work for the same price even though the client was happy to spend more on the overall project. This may account for some or in the best case, all, of their overspent time, although again a savvy client may turn to other consultants that agree to omit variation opportunities.

@@

@Lovallo2003 called the practice of analysing a project based on project details without considering unforeseen risks the 'inside view'. A survey by @Moores1992 aligned with this theory and found the most popular response for the supposed reason for cost overruns was over-optimistic estimates (51%). This contrasts with an 'outside view' where a subject is analysed by comparison to other subjects and disregarding specific subject details. For example, research has shown that if people make predictions about their skills after being exposed to an 'outside view' (a summary of other people's skills), their predictions are significantly more accurate. According to @Lovallo2003, this 'outside view' can be applied to complex projects, and stipulates that the details of the current project should be ignored in favour of analysing outcomes of several similar projects. Mobilising the outside view in this way is called reference class forecasting [@Lovallo2003]. One method @Lovallo2003 recommended was to obtain correlation statistics from past similar projects - the correlation between the forecast cost and the actual cost. The correlation for the current project can then be estimated via a statistical model, which is then used to adjust the forecast cost made by detailed analysis (the 'inside view') [@Flyvbjerg2011]. As demonstrated in the surveys, @Flyvbjerg2011 found that managers rarely analysed data from similar projects even though it could significantly improve cost predictions. This discovery led the team to formally introduce 'outside view' methods into some institutions.

@@

![This model was implemented by the British government for a rail project that is still under completion [@Flyvbjerg2011]](images/lit_review/UKrail.JPG)

@@

Despite the predictive success in statistical and machine learning cost estimation models as well as the revealing outcomes from 'inside' and 'outside' view studies, reference class forecasting there has been a lack of industry uptake. Two reports reviewing cost estimation in the software industry investigated reasons why data-driven forecasting models have not been translated to industry. The key causes were:

@@

Reports on other industries were not available, however the insights are general and may explain the lack of adoption in other industries (such as the construction industry).

In the software industry it is also possible the problem of fixed price contract blowouts has been solved by an alternative contract structure, which has made cost estimation models unnecessary. In contrast to construction, IT projects are a relatively new practice and the industry has had success in trialing alternative contractual arrangements. Through experimentation, the Agile movement was developed. In summary, the method treats both cost and time as fixed quantities for a project, and any change or additional work can only be accommodated if another, less important requirement is excluded [@Badenfelt2011]. Furthermore, the project outcomes are continually adjusted and revised based on frequent progress review points that assess unexpected problems. The flexible upfront arrangement with clients has often been found to forge better long-term relationships [Cockburn2001]. This agile methodology suits the nature of software development work, where curtailing the scope of work is relatively straightforward. Also, since the industry is more modern, the new methodology was able to sweep across the field and could be proven to deliver better budget and product outcomes to clients [@Rasmusson2010].

Unfortunately, in other industries such as the construction industry, infrastructure and building projects are steeped in a traditional process, where contractual norms have developed over a much longer time period [@Badenfelt2011]. The industry is so large, broad and well established that it has inertia in the way projects are estimated, and contractually bound. Furthermore, the industry would struggle to take on an agile approach as the scope cannot be easily reduced as the project unfolds - the project must result in construction and building certification which implies a high-level of legal responsibility and due diligence.

This project aims to assist a company in the construction industry in managing problematic fixed price projects, since fixed price contracts are unavoidable. The blocks preventing industry adoption, such as providing a clear decision framework and incorporating expert judgment will be carefully addressed.

@@

Cost estimation is currently performed in the case study company by first carefully reviewing preliminary drawings where time costs for each task are estimated and summed by an experienced manager (the 'inside view'). This is often crosschecked with a value based off a percentage of the estimated final cost of construction (generally the client budget). No formal mathematical method of comparison to similar projects is performed. This is the case for a number of possible reasons:

@@

The company had a wealth of data but limited means to extract insight. This study exploits the CRM data by performing more sophisticated statistical analysis with the intention of building a predictive model to improve cost estimation. Thousands of past projects are available for analysis, as opposed to previous studies that used on average 15-20 cases and at most 300 [@Finnie1997; @Pai2013; @Shin2015]. This improves the potential for accurate predictions. If successful, the model also has higher potential for managerial uptake as the algorithm will be directly built and trained on internal company data. Managers could interrogate how the model outputs were calculated and relate to the actual projects used in the model or speak to a colleague that was involved. This may give the presented 'outside view' a stronger meaning and more potential to influence the final fee proposal.

@@

The database also excludes detailed technical information about each project. Such details are found in the preliminary drawings describing the project (square meterage, number of storeys), as well as project summary documents provided by the client to enable a quote. However, this level of information actually contradicts the purpose of the 'outside view', the intention of which is to present similar projects without being over-influenced by finer details. Given the lack of detailed information describing the physical size of projects, it was not reasonable for an algorithm to predict the numeric fee. Instead, it was more logical to predict the profitability of a project i.e. 'return per dollar'. This measure represents the ratio of the amount invoiced to business costs, and can easily be compared between projects of different sizes.

@@

$Overall\ Profit = Revenue\ from\ Project\ Invoices\ - (Employee + Business\ Costs)$

@@

In comparison to previous studies, this project advances the body of work on cost estimation in three ways: the data set is larger than other studies and sourced from a single company's CRM database, ensemble tree methods and model blending was tested, and the prediction results were integrated into a decision framework for the business. These points are outlined in more detail below.

In past research, cost estimation data sets generally consisted of dozens of projects from a range of a companies and even countries. This thesis case study only had access to the company's internal data but this contained over 2,000 past jobs - far more than the usual cost estimation study and with many more variables. Furthermore, the problem of cost estimation is applied to a consulting company in the construction industry, which differs from other studies in the construction industry that estimate actual construction costs. Consulting in the construction industry can be likened more to software development effort estimation although the nature of the work differs considerably.

@@

The next chapter, Chapter 3 is a thorough review of the literature to date that relates to this thesis. This covers work solving the cost-estimation problem, studies on statistical and machine-learning models in business applications, and a review of potential statistical and machine learning methods.

@@

Researching outcomes of comparable past studies may impart how to best approach the problem and reveal shortcomings that can be improved upon. This chapter provides an overview of the literature describing early and current cost estimation methods. The most prominent machine learning and statistical methods shall then be reviewed followed by a summary of papers on more general business applications that have tried a wider variety of methods. Gaps in past research will also be highlighted.

@@

A notable amount of literature tested the predictive power of Multiple Linear Regression in the construction cost estimation problem. Often, Linear Regression was the only model assessed, without comparison to other methods such as Neural Networks, which first started appearing in literature in the 1990's [@Kim2004]. However, even post-1990, many studies focused on Linear Regression only. This may be because the technique is straightforward, easy to use and widely available in statistical packages [@Chan2005]. Some studies showed that Neural Networks outperform Regression, however other studies established they are approximately equal [@Kim2004; @Attalla2003]. @Dissanayaka1999 suggested that Regression models could be performed first to determine variable importance and condense the number of variables, then Neural Networks could be used for better estimation. Boosted Trees may perform as well as Neural Networks while providing even more insight than Regression models because it is possible to produce partial dependency plots. These visualise the relationship between each variable and the response variable (cost).

@@

1. Storing a collection of projects with key variable values
2. Once a new project or case arrives, similar cases are retrieved. This can be achieved either algorithmically with a distance function applied to the variables, or manually by users reviewing past cases. The Nearest Neighbour algorithm has been used to find the distance between cases in previous studies. It is programmed to calculate the Euclidean distance in n-dimensional space between cases, where each variable is a dimension [@Kumar2007].
3. The cost of the new project is estimated by extrapolating characteristics of the similar cases to the new case. An experienced decision maker can also do this either algorithmically or manually.

@@

Similar to the construction industry, expert judgment or detailed analysis, is the most widely practised method for effort estimation [@Shepperd1996; @Molokken2003]. This is despite many years of research being dedicated to developing algorithmic models that, in the research context, outperform expert judgment. A survey by @Moores1992 found that software tools assisting a software project were in significant use, but only the project planning tools, not the cost estimating tools. This suggests a mental link for project managers to price jobs based on *details* as opposed to an outside view. Another study by @Heemstra1992 found there was no evidence estimation accuracy improved when estimation tools were used. @Bergeron1992 similarly found jobs that used algorithmic models were actually associated with less accurate estimates. However, this trend may be coincidental due to a lack of cases where estimation tools were used [@Molokken2003]. The scarcity of real project evidence that algorithmic estimation tools improve estimations and an unintuitive mental jump could be reasons contributing to detailed expert analysis remaining the most widespread technique and mirrors what was found in the construction industry.

@@

This section assesses the details and performance of the algorithmic models, important variables, and gaps in the literature.

@@

Effort estimation in the software industry has followed a similar path to cost prediction of construction projects. Expert judgment via detailed analysis is still the predominant method for effort/cost estimation. Case based reasoning and algorithmic methods such as Neural Networks have found some success in effort estimation (at similar levels) but have not been successfully adopted by industry. It was again noted that CBR and visibility of model structure resonated more with decision makers. The software industry is unique in that agile methods of delivery are changing the contractual approach of consulting [@Badenfelt2011]. This innovative method keeps delivery times short and defines projects as something with a fixed cost and time. If a change occurs, other less important elements are discarded to maintain the fixed amount. Software development is a relatively new field in comparison to construction, where the tradition of fixed fee structures is difficult to reinvent. Also, portions of construction projects cannot be simply curtailed.

There is a need for research on algorithmic models that predict effort (and not construction cost) for consulting companies in the construction industry. From findings in previous literature, the models should reveal their predictive structure to decision makers and be based on internal company data. Review of the literature revealed a need for an industry focused decision framework that outlines how to integrate the predictive model in a way that optimises profits.

@@

@@

If structured correctly, Linear Regression results can show the importance of the explanatory variables. The variables must be first standardised to one another (i.e. scaled to the same mean and standard deviation). Then, their coefficients can be directly compared [@Putler2012]. The magnitude of the coefficients (the slopes) indicates how *much* a variable changes with the response variable. However, if two or more predictor variables are correlated, the values of their coefficients are unstable. More specifically, this means that if a model has two correlated variables, and is re-run with a slightly different set of data, the coefficients for the two variables may be quite different. This is because explanatory variables are assumed to be independent, and if variables are correlated, it breaks this assumption. On the other hand, correlated variables do not impact the predictive accuracy of the model [@Putler2012].

In linear regression, no assumptions are made about the distributions of the variables. Skewed explanatory variable distributions are acceptable but it is generally better to normalise them by, for example, taking the logarithm or square root of the variable. The reason for this is that the more extreme values from the skewed distribution may create high leverage points and influence the slope (variable coefficients) too strongly [@Chatterjee1986]. The distribution of the residuals on the other hand, should always be Gaussian with a mean of 0. This is because if modeled well, the regression equation should follow the shape of the data [@Hastie2009].

@@

Nevertheless, the theory behind Linear Regression can be adapted to the prediction of probabilities pertaining to a binary response variable (represented by 1 or 0). First, a sigmoid function is assumed as the functional form of the probabilities against an example variable X. This is because it is pragmatic and works well: sigmoid functions can be bound by 0 and 1, and a logit transformation of the probability results in a linear relationship between the log odds and X. Refer to the equation below where *p* represents probability [@Macdonald1975]:

@@

Because of the linear relationship, an equation can be fit to the log odds of the binary response variable against each of the explanatory variables using linear regression. Then, the linear equation describing the log odds can be transformed back to probability by taking the inverse log, i.e. the exponential and rearranging [@Macdonald1975]:

@@

@@

![Plot of log odds vs. an 'x' variable and probability vs. the 'x' variable. To convert log odds to probability, an exponential transformation is first made followed by the transformation from odds to probability [@Lowry2016].](images/lit_review/logreg.JPG)

@@

In one case, this classifier succeeded because of its ability to be refined incrementally with each piece of new data, as opposed to re-calibrating the entire model. Naive Bayes was used in a complex spam detection system where new spam emails or toxic text themes could be quickly added to the filtering model. Although the given example is different to cost estimation, the Naive Bayes method generally provides a good benchmark to compare against more complex models that should outperform it [@Caruana2006].

@@

Decision trees have a different theoretical foundation to traditional prediction methods such as regression. They are created by a progression of binary splits of the data set that attempt to group similar response variable values. Before a split is made, each explanatory variable and each value within the explanatory variables is tested for how 'purely' it partitions the response variable into two subsets. For example, the explanatory variable 'gender' might split a response variable 'hair length' of a class of students into two subsets of students with more similar hair lengths. This similarity is called 'purity' and can be measured using several formulas. An example for regression (a problem with a numeric response variable) is [@Hastie2009]:

@@

$$
\begin{aligned}
SSN\ \textrm{is the sum of squared errors of a node (dataset before the split)} \\
y_i\ \textrm{is the}\ i \textrm{th response variable value} \\
\bar{y}\ \textrm{is the mean response variable value in the node} \\
SSL\ \textrm{is the sum of squared errors in the left dataset that resulted from the split} \\ 
SSR\ \textrm{is the sum of squared errors in the right dataset that resulted from the split} 
\end{aligned}
$$

@@

![Simple 4-Node Bayesian Network](images/lit_review/BN_1.png)

@@

![4-Node Bayesian Network with conditional independencies](images/lit_review/BN_2.png)

@@

In this way, Bayesian Networks reduce computations required to find the probability of a unique combination given other explanatory variable values while still taking into account many conditional dependencies that Naive Bayes, for example, cannot [@Barber2012]. Also, some equations for a combination are now clones of another combination and do not require recalculation. Calculating the conditional probability tables for each variable is still much more computationally expensive than Naive Bayes (as Naive Bayes is the simplest form of a Bayesian Network) and variables with many categories dramatically increase the expense [@Zhang2004].

@@

Algorithmic Neural Networks were inspired by Neural Networks of the human nervous system. In the human body, biological neurons receive electro-chemical input into their dendrites from the terminals of adjacent neurons. If enough electro-chemical input is received, the neuron will 'fire', which means a signal will be sent from the dendrite end, through the axon to its terminal. Its terminal then sends signals to adjacent dendrites of other neurons and the process propagates. Neurons are able to learn over time by strengthening the connections of adjacent neurons that frequently cause the next one to fire. This means the required firing threshold for the receiving neuron lowers in strengthened connections [@Putler2012].

@@

The method for finding the boundary, however, is different to Logistic Regression. The aim is to find the hyperplane with the widest clear margin between data points from the two categories, while allowing a certain number of misclassifications (this is usually optimised for the problem). To do this, the equation for the hyperplane is expressed as follows [@Fletcher2009]:

@@

The plane is expressed with vectors because they can be defined such that $\boldsymbol{w}$ is perpendicular to $\boldsymbol{x}$. This aids optimisation calculations to find the hyperplane with the widest margin, i.e. two hyperplanes with the largest perpendicular distance between them that contains less than the allowable number of misclassified points [@Fletcher2009]. The plot below demonstrates this concept and has no misclassifications:

@@

Advantages of SVM's include the ability to model nonlinear relationships in data, and that no assumptions about the distributions of the data must be made as the model is non-probabilistic. They are also known to be robust over different samples and perform well in high-dimensional feature space [@Auria2008]. On the other hand, the output is in the form of distance to the boundary (as opposed to probability) and results are not transparent (similar to Neural Networks) [@Caruana2006]. Also, SVM's can be very slow to train and therefore not suitable for industry purposes [@Auria2008]. Because SVM's perform well in high-dimensional space and can form non-linear rules they have been applied to protein classification (medicine) and text and image recognition [@Byun2002].

@@

### Gap

>>>>

Engaging in loss making jobs for fixed fees is a major problem in consulting, particularly in the competitive construction industry. This thesis investigates whether machine learning techniques applied to a company's passively collected internal data could help avoid loss making jobs or help tactfully choose when to enforce stricter contracts. It was found that in a specific decision framework, a case study's profits could be improved 9% by avoiding approximately 4% of projects. Alternative decision frameworks are also proposed and evaluated. The internal data collected by the case study company described each of their projects in terms of employee positions, fraction of project time completed by each employee, client characteristics, project characteristics (time span, hours, technical details) and invoicing history. Algorithmic methods such as Logistic Regression, Random Forests, Boosted Trees, Naive Bayes, and Bayesian Networks were applied as well as blended combinations of these methods. A decision scenario that rejected projects above a sequence of tested thresholds was run in order to find the optimal threshold for profit improvements. This process was repeated on hundreds of models built from different training subsets of the data in order to obtain a 95% statistical confidence interval of profit improvements. The blended Logistic Regression model outperformed other methods and produced a 95% confidence interval of 6.5 - 11.5% profit improvements. Further work is recommended to test user input estimation and user interface designs. The findings from this research have the potential to assist managers in reducing losses by highlighting risky projects and guiding project-based changes to fee structures.

@@

**ANOVA** - Analysis of variance. A statistical test that can be applied to find whether there is a significant difference between the means of more than two groups.

@@

**B2B** - Business to business. Commerce transactions between two businesses.

**BIM** - Building information modeling. A digital representation of physical characteristics of an object or structure.

@@

**CLV** - Customer lifetime value. Prediction of the profit a customer will bring over the course of a business' relationship with the customer.

**CRM** - Customer relationship management. Refers to a common type of software used by businesses to record client and project details as well as employee time sheet records. Generally accessible to each employee in a company.

@@

**FNR** - False negative rate. The proportion of negatives that were incorrectly identified (false negatives).

@@

**ROC** - Receiver operating characteristic. An ROC curve plots the performance of a binary classification algorithm its scoring threshold is varied.

@@

**TNR** - True negative rate. The proportion of negatives that were correctly identified (true negatives).

@@

Predicting how long it will take a person to do something is a notoriously tricky task. People make mental calculations of this sort every day - estimating how long it will take to drive to meet someone on time, how many tasks you can check off in a day, or how long it will take to cook a dish. As most people have experienced, these estimates can be particularly inaccurate if the task is new and difficult. These errors are trivial in everyday activities, but for consulting businesses that solve complex problems, similar time estimation errors determine their financial well being.

@@

“It’s tough to make predictions, especially about the future” - Danish Proverb

@@

Another strategy for managing this risk is by stipulating a specific one-off variation opportunity in the contract. This applies to consulting work that falls under the client's larger project budget. The variation would specify a percentage fee of the overall project cost that can overrule the original fixed price. The purpose of this rule is to guard against the case where the project creeps in size, and the consultant finds themselves gradually undertaking more work and more work for the same price even though the client was happy to spend more on the overall project. This may account for some, or in the best case all, of their overspent time, although again a savvy client may turn to other consultants that agree to omit variation opportunities.

@@

@Lovallo2003 called the practice of analysing a project based on project details without considering unforeseen risks the 'inside view'. A survey by @Moores1992 aligned with this theory and found the most popular response for the supposed reason for cost overruns was over-optimistic estimates (51%). This contrasts with an 'outside view' where a subject is analysed by comparing it to other subjects and disregarding specific subject details. For example, research has shown that if people make predictions about their skills after being exposed to an 'outside view' (a summary of other people's skills), their predictions are significantly more accurate [@Lovallo2003]. This 'outside view' can be applied to complex projects, and stipulates that the details of the current project should be ignored in favour of analysing outcomes of several similar projects. Mobilising the outside view in this way is called reference class forecasting [@Lovallo2003]. One method @Lovallo2003 recommended was to obtain correlation statistics from past similar projects - the correlation between the forecast cost and the actual cost. The correlation for the current project can then be estimated via a statistical model, which is then used to adjust the forecast cost made by detailed analysis (the 'inside view') [@Flyvbjerg2011]. As demonstrated in the surveys, @Flyvbjerg2011 found that managers rarely analysed data from similar projects even though it could significantly improve cost predictions. This discovery led the team to formally introduce 'outside view' methods into some institutions.

@@

![This model was implemented by the British government for a rail project that was under completion at the time of publication [@Flyvbjerg2011]](images/lit_review/UKrail.JPG)

@@

Despite the predictive success in statistical and machine learning cost estimation models as well as the revealing outcomes from 'inside' and 'outside' view studies, there has been a lack of industry uptake. Two reports reviewing cost estimation in the software industry investigated reasons why data-driven forecasting models have not been translated to industry. The key causes were:

@@

Reports on other industries were unavailable, however the insights are general and may explain the lack of adoption in other industries (such as the construction industry).

In the software industry it is also possible the problem of fixed price contract blowouts has been solved by an alternative contract structure, which has made cost estimation models unnecessary. In contrast to construction, IT projects are a relatively new practice and the industry has had success in trialing alternative contractual arrangements. Through experimentation, the Agile movement was developed. The method treats both cost and time as fixed quantities for a project, and any change or additional work can only be accommodated if another, less important requirement is excluded [@Badenfelt2011]. Furthermore, the project outcomes are continually adjusted and revised based on frequent progress review points that assess unexpected problems. The flexible upfront arrangement with clients has often been found to forge better long-term relationships [Cockburn2001]. This agile methodology suits the nature of software development work, where curtailing the scope of work is relatively straightforward. Also, since the industry is relatively young, the new methodology was able to sweep across the field and was shown to deliver better budget and product outcomes to clients [@Rasmusson2010].

Unfortunately, in other industries such as the construction industry, infrastructure and building projects are steeped in a traditional process, where contractual norms have developed over a much longer time period [@Badenfelt2011]. The industry is so large, broad and well established that it has inertia in the way projects are estimated, and contractually bound. Furthermore, the industry would struggle to take on an agile approach as the scope cannot be easily reduced as the project unfolds - the project must result in construction and building certification which implies a high-level of time, legal responsibility and due diligence.

This project aims to assist a company in the construction industry in managing problematic fixed price projects, since fixed price contracts can seem unavoidable. The blocks preventing industry adoption, such as providing a clear decision framework for how to implement the model will be carefully addressed.

@@

Cost estimation is currently performed in the case study company by first carefully reviewing preliminary drawings where time costs for each task are estimated and summed by an experienced manager (the 'inside view'). This is often crosschecked with a value that is based off a percentage of the estimated final cost of construction (generally the client budget). No formal mathematical method of comparison to similar projects is performed. This is the case for a number of possible reasons:

@@

The company had a wealth of data but limited means to extract insight. This study exploits the CRM data by performing more sophisticated statistical analysis with the intention of building a predictive model to improve cost estimation. Thousands of past projects are available for analysis, as opposed to previous studies that used on average 15-20 cases and at most 300 [@Finnie1997; @Pai2013; @Shin2015]. This improves the potential for accurate predictions. If successful, the model also has higher potential for managerial uptake as the algorithm will be directly built and trained on internal company data. Managers could interrogate how the model outputs were calculated and relate to the actual projects used in the model or speak to a colleague that was involved. This may make the presented 'outside view' more relatable and have more potential to influence the final fee proposal.

@@

The database also excludes detailed technical information about each project. Such details are found in the preliminary drawings describing the project (square meterage, number of storeys), as well as project summary documents provided by the client to enable a quote. However, this level of information actually contradicts the purpose of the 'outside view', the intention of which is to present similar projects without being over-influenced by finer details. Given the lack of detailed information describing the physical size of projects, it was not reasonable for an algorithm to predict the numeric fee. Instead, it was more logical to predict the profitability of a project i.e. 'return per dollar'. This measure represents the ratio of the project profit to business costs, and can easily be compared between projects of different sizes.

@@

$Overall\ Profit = Revenue\ from\ Project\ Invoices\ - (Employee\ Costs + Business\ Costs)$

@@

In comparison to previous studies, this project advances the body of work on cost estimation in three ways: the data set is larger than other studies and sourced from a single company's CRM database, ensemble tree methods and model blending were tested, and the prediction results were integrated into a decision framework for the business. These points are outlined in more detail below.

In past research, cost estimation data sets generally consisted of dozens of projects from a range of a companies and even countries. This thesis' case study only had access to the company's internal data but this contained over 2,000 past jobs - far more than the usual cost estimation study and with many more variables. Furthermore, the problem of cost estimation is applied to a consulting company in the construction industry, which differs from other studies in the construction industry that estimate actual construction costs. Consulting in the construction industry can be likened more to software development effort estimation although the nature of the work differs considerably.

@@

The next chapter, Chapter 3, is a thorough review of the literature to date that relates to this thesis. This covers work solving the cost-estimation problem, studies on statistical and machine-learning models in business applications, and a review of potential statistical and machine learning methods.

@@

Assessing outcomes of comparable past studies may impart how to best approach the problem and reveal shortcomings that can be improved upon. This chapter provides an overview of the literature describing early and current cost estimation methods. The most prominent machine learning and statistical methods shall then be reviewed followed by a summary of papers on more general business applications that have tried a wider variety of methods. Gaps in past research will also be highlighted.

@@

A notable amount of literature tested the predictive power of Multiple Linear Regression in the construction cost estimation problem. Often, Linear Regression was the only model assessed, without comparison to other methods such as Neural Networks, which first started appearing in literature in the 1990's [@Kim2004]. However, even post-1990, many studies focused on Linear Regression only. This may be because the technique is straightforward, easy to use and widely available in statistical packages [@Chan2005]. Some studies showed that Neural Networks outperform Regression, however other studies established they are approximately equal [@Kim2004; @Attalla2003]. @Dissanayaka1999 suggested that Regression models could be performed first to determine variable importance and condense the number of variables, and then Neural Networks could be used for better estimation. Boosted Trees may perform as well as Neural Networks while providing even more insight than Regression models because it is possible to produce partial dependency plots. These visualise the relationship between each variable and the response variable (cost).

@@

1. Storing a collection of projects with key variable values
2. Once a new project or case arrives, similar cases are retrieved. This can be achieved either algorithmically with a distance function applied to the variables, or manually by users reviewing past cases. The nearest neighbour algorithm has been used to find the distance between cases in previous studies. It is programmed to calculate the Euclidean distance in n-dimensional space between cases, where each variable is a dimension [@Kumar2007].
3. The cost of the new project is estimated by extrapolating characteristics of the similar cases to the new case. An experienced decision maker can also do this either algorithmically or manually.

@@

Similar to the construction industry, expert judgment or detailed analysis, is the most widely practised method for effort estimation [@Shepperd1996; @Molokken2003]. This is despite many years of research being dedicated to developing algorithmic models that, in the research context, outperform expert judgment. Another study by @Heemstra1992 found there was no evidence estimation accuracy improved when estimation tools were used. @Bergeron1992 similarly found jobs that used algorithmic models were actually associated with less accurate estimates. However, this trend may be coincidental due to a lack of cases where estimation tools were used [@Molokken2003]. The scarcity of real project evidence that algorithmic estimation tools improve estimations and an unintuitive mental jump could be reasons contributing to detailed expert analysis remaining the most widespread technique and mirrors what was found in the construction industry.

@@

This section assesses the details and performance of the algorithmic models in the IT industry, including their important variables, and gaps in the literature.

@@

Effort estimation in the software industry has followed a similar path to cost prediction of construction projects. Expert judgment via detailed analysis is still the predominant method for effort/cost estimation. Case based reasoning and algorithmic methods such as Neural Networks have found some success in effort estimation (at similar levels) but have not been successfully adopted by industry. It was again noted that CBR and visibility of model structure resonated more with decision makers. The software industry is unique in that agile methods of delivery are changing the contractual approach of consulting [@Badenfelt2011].

There is a need for research into algorithmic models that predict effort (and not construction cost) for consulting companies in the construction industry. For industry adoption, the models should reveal their predictive structure to decision makers and provide a decision framework that outlines how to integrate the predictive model in a way that optimises profits.

@@

$$
\begin{aligned}
\beta\ \textrm{represents the set of coefficients for a model} \\
N\ \textrm{is the number of points in a data set} \\
y_{i}\ \textrm{is the response variable value for the}\ i\textrm{th case in the data set} \\
f(x_{i})\ \textrm{is the predicted response variable value for the}\ i\textrm{th case in the data set}
\end{aligned}
$$

@@

If structured correctly, Linear Regression results can show the importance of the explanatory variables. The variables must be first standardised to one another (i.e. scaled to the same mean and standard deviation). Then, their coefficients can be directly compared [@Putler2012]. The magnitude of the coefficients (the slopes) indicate how *much* a variable changes with the response variable. However, if two or more predictor variables are correlated, the values of their coefficients are unstable. More specifically, this means that if the a model has two correlated variables, and is re-run with a slightly different set of data, the coefficients for the two variables may be quite different. This is because explanatory variables are assumed to be independent, and if variables are correlated, it breaks this assumption. On the other hand, correlated variables do not impact the predictive accuracy of the model [@Putler2012].

In Linear Regression, no assumptions are made about the distributions of the variables. Skewed explanatory variable distributions are acceptable but it is generally better to normalise them by, for example, taking the logarithm or square root of the variable. The reason for this is that the more extreme values from the skewed distribution may create high leverage points and influence the slope (variable coefficients) too strongly [@Chatterjee1986]. The distribution of the residuals on the other hand, should always be Gaussian with a mean of 0. This is because if modeled well, the regression equation should follow the shape of the data [@Hastie2009].

@@

Nevertheless, the theory behind Linear Regression can be adapted to the prediction of probabilities pertaining to a binary response variable. First, a sigmoid function is assumed as the functional form of the probabilities against an example variable X. This is because it is pragmatic and works well: sigmoid functions can be bound by 0 and 1, and a logit transformation of the probability results in a linear relationship between the log odds and X. Refer to the equation below where *p* represents probability:

@@

[@Macdonald1975]

Because of the linear relationship, an equation can be fit to the log odds of the binary response variable against each of the explanatory variables using linear regression. Then, the linear equation describing the log odds can be transformed back to probability by taking the inverse log, i.e. the exponential and rearranging:

@@

[@Macdonald1975]

@@

![Plot of log odds vs. an 'X' variable and probability vs. the 'X' variable. To convert log odds to probability, an exponential transformation is first made followed by the transformation from odds to probability [@Lowry2016].](images/lit_review/logreg.JPG)

@@

In one industry application, this classifier succeeded because of its ability to be refined incrementally with each piece of new data, as opposed to re-calibrating the entire model. Naive Bayes was used in a complex spam detection system where new spam emails or toxic text themes could be quickly added to the filtering model. Although the given example is different to cost estimation, the Naive Bayes method generally provides a good benchmark to compare against more complex models that should outperform it [@Caruana2006].

@@

Decision trees have a different theoretical foundation to traditional prediction methods such as regression. They are created by a progression of binary splits of the data set that attempt to group similar response variable values. Before a split is made, each explanatory variable and each value within the explanatory variables is tested for how 'purely' it partitions the response variable into two subsets. For example, the explanatory variable 'gender' might split a response variable 'hair length' of a class of students into two subsets of students with more similar hair lengths. This similarity is called 'purity' and can be measured using several formulas. An example for regression (a problem with a numeric response variable) is:

@@

$$
\begin{aligned}
SSN\ \textrm{is the sum of squared errors of a node (dataset before the split)} \\
SSL\ \textrm{is the sum of squared errors in the left dataset that resulted from the split} \\ 
SSR\ \textrm{is the sum of squared errors in the right dataset that resulted from the split} \\
y_i\ \textrm{is the}\ i \textrm{th response variable value} \\
\bar{y}\ \textrm{is the mean response variable value in the node}
\end{aligned}
$$

@@

![Simple 4 node Bayesian Network](images/lit_review/BN_1.png)

@@

![4 node Bayesian Network with conditional independencies](images/lit_review/BN_2.png)

@@

In this way, Bayesian Networks reduce computations required to find the probability of a unique combination given other explanatory variable values, while still taking into account many conditional dependencies that Naive Bayes, for example, cannot [@Barber2012]. Also, some equations for a combination are now clones of another combination and do not require recalculation. Calculating the conditional probability tables for each variable is still much more computationally expensive than Naive Bayes (as Naive Bayes is the simplest form of a Bayesian Network) and variables with many categories dramatically increases the expense [@Zhang2004].

@@

Neural Networks of the human nervous system inspired Algorithmic Neural Networks. In the human body, biological neurons receive electro-chemical input into their dendrites from the terminals of adjacent neurons. If enough electro-chemical input is received, the neuron will 'fire', which means a signal will be sent from the dendrite end, through the axon to its terminal. Its terminal then sends signals to adjacent dendrites of other neurons and the process propagates. Neurons are able to learn over time by strengthening the connections of adjacent neurons that frequently cause the next one to fire. This means the required firing threshold for the receiving neuron lowers in strengthened connections [@Putler2012].

@@

The method for finding the boundary, however, is different to Logistic Regression. The aim is to find the hyperplane with the widest clear margin between data points from the two categories, while allowing a certain number of misclassifications (this is usually optimised for the problem). To do this, the equation for the hyperplane is expressed as follows:

@@

The plane is expressed with vectors because they can be defined such that $\boldsymbol{w}$ is perpendicular to $\boldsymbol{x}$. This aids in optimisation calculations to find the hyperplane with the widest margin, i.e. two hyperplanes with the largest perpendicular distance between them that contains less than the allowable number of misclassified points [@Fletcher2009]. The plot below demonstrates this concept and has no misclassifications:

@@

Advantages of SVM's include the ability to model nonlinear relationships in data, and that no assumptions about the distributions of the data must be made as the model is non-probabilistic. They are also known to be robust over different samples and perform well in high-dimensional feature space [@Auria2008]. On the other hand, the output is in the form of distance to the boundary (as opposed to probability) and results are not transparent (similar to Neural Networks) [@Caruana2006]. Also, SVM's can be very slow to train and sometimes not suitable for industry purposes [@Auria2008]. Because SVM's perform well in high-dimensional space and can form non-linear rules they have been applied to protein classification (medicine) and text and image recognition [@Byun2002].

@@

### Research Gap
